The not so widening achievement gap: an interntational comparison of the evolution of the SES achievement gap from 2000-2015

### 2017-07-14 paper/paper.Rnw
Redid mistake from stan branch

<<default_conf>>=
  library(knitr)
  library(arm)
  library(saves)
  library(haven)
  library(PISA2000lite)
  library(PISA2003lite)
  library(PISA2006lite)
  library(PISA2009lite)
  library(PISA2012lite)
  library(intsvy)
  library(cimentadaj) # devtools::install_github("cimentadaj/cimentadaj")
  library(countrycode) # For region variable
  library(car)
  library(SAScii)
  library(inequalityintsvy) # devtools::install_github("cimentadaj/inequalityintsvy")
  library(lme4)
  library(modelr)
  library(tidyverse)
  library(ggrepel)
  library(xtable)
  library(ggthemes)

@@

      # Calculate median math score of all PV's
      .x$dv <- apply(.x[test_vars], 1, median, na.rm = T)

@@

\section{Literature Review}
% You have a bunch of citations to take from here
Educational inequality and its long term impacts is a topic that has been prominent in the social science literature for the past 30 years. The idea of meritocracy and intergenerational transfers have motivated, in conservative terms, much of the research on social mobility and social inequalities, for a good part of the 20th century and 21st century. When James Coleman released his famous Coleman report \citep{coleman1966}, it raised the topic to starlight by suggesting that family SES and student performance are tightly linked. The topic has been studied extensively since the report was released and several authors have contested whether the relationship is an invariable social law or a product of institutional arrangements. For a detailed review of the long literature on educational inequality, please see \citet{gamoran2001}. But fast-forward until today, we've understood the relationship much better.

The work of James Heckman, for example, has clearly showed that cognitive and non-cognitive inequalities are present before entering school \citep{heckman2006}. This means that inequality starts strong before the schooling of a child begins and schools can do so much to close it. These inequalities can be narrowed significantly with high quality learning environments, specially for children coming from unfavourable conditions. But one of the most important findings from \citet{cunha2006} is that the cognitive level of a child in time \(t\) is a direct function of the experiencies from time \(t-1\). Although this sounds logical, it's importance is not understood entirely.

The implications of this cumulative model suggests that it is much less cost-effective to invest in children in time \(t\) that in \(t-1\). That is why the whole debate about narrowing inequality between high and low SES families should start in early education -- it is the cheaper than investing in later education and it brings about the largest returns. Despite all of these findings and investments to reduce the gaps, we still find that the relationship between parental education or socio-economic status is present in virtually all empirical studies of social mobility and inequalities \citep{breen1997, breen2007, waldfogel2006, bradbury2015, chetty2016}. What we don't know is how and why this relationship is formed from such an early age. Moreover, many have suggested different explanations and the topic has been very contested \citep{brooks1997}.

@@

From Denmark, for example, the literature has learned a great deal about how to improve social mobility. But that was possible by first learning that Denmark is by far one of the most mobile countries, boosting the chances of upward mobility specially for less advantaged children \citep{bjorklund2009, jaeger2007}. The important finding came when we discovered the main reason behind their social escalator: it's educational system. For example, research by \citet{esping_waldfogel2012} and \citet{bauchmuller2014}, show that the Danish early education system has important and longstanding impacts in helping children reach their fullest potential. It does so very strongly for children which otherwise would be doomed to have a bleak future. Moreover, it tracks students into different curricula very late compared to other Europen countries (age 16), something which has been linked to less educational inequality. These two traits make the Danish system uniquely effective. The early schooling experience attempts to level children to the same degree and this process is not stopped by different curricular tracks because it starts at around age 16 when cognitive abilities are less malleable. In short, the importance of this finding, is that we should step back and first study whether the effect is there, how big is it, and then proceed to find the causes behind it.

The first attempt to study the evolution of the achievement gap has found that the gap in cognitive abilities between High-SES and Low-SES kids has been widening over the years. The literature on the topic has mainly concentrated on studying the American case \citep{reardon2011} but other international evidence is emerging with a very similar landscape. The United States is usually the case of study because it is the only country where cognitive testing is present as early as 1940. This time-trend has allowed researchers to study achievement gaps in a lengthy period of time. Using this information, \citet{reardon2011} is the first to investigate the evolution of the cognitive gap and the results are very surprising. Not only has the cognitive gap between the 90th income percentile and the 10th income percentile grown over time, but it has grown faster and to be wider than the highly contested white-black gap \citep{magnuson2008}. The gaps have actually reversed and now a days we find that the income achievement gap is nearly twice as large as the black-white achievement gap (quite the opposite to 20 years back).

\citet{reardon2011} finds that the increase in the gaps has occured predominantely from the 1970's until the 2000's. In fact, the hard numbers suggest that the gap widened by 40-50\%. The author also estimates the rate of change using data as early as 1940 and finds an even higher increase of   75\%. Given that the studies before 1970 are less reliable in terms of comparability and sampling design, the author computes all results for both before/after 1970. To provide a definitive answer to the size of the gap, \citet{reardon2011} concludes that the U.S gap between the 90th and 10th income percentile is at about 1.25 standard deviations in the year 2001. Using longitudinal data, \citet{bradbury2015} find similar results. Their empirical analysis suggests that for American 14 year olds, the gap is above 1 standard deviation but lower than 1.25. Surprisingly, \citet{duncan2011} find similarly to the previous studies and confirm a gap of  1.25-1.50 standard deviations. One important drawback of these studies is that they don't present the uncertainty of this estimates. Not necessarily to gauge their statistical significance, but to simply asses how much we can trust their accuracy. It could be that the gap is at  1.25 standard deviations, varying up to 1.75 and down to 0.80. Right now we don't know the upper/lower bound of this estimate, making it difficult to compare when new evidence arises.

Interestingly, this widening of the achievement gap has been paralleled by a growth of income inequality, a very suggestive explanation. However likely, this is a highly contested topic. A recent \href{http://www.epi.org/files/pdf/101972.pdf}{report} by the Economic Policy Institute finds that the black-white income gap has been growing since 1979, very similar to the increase of income inequality. However, if the achiement gap would be a function of income inequality, we should expect for the white-black achievement gap to be widening rather than narrowing, and those are not the findings given by \citet{reardon2011} and \citet{magnuson2008}.

\footnote{The black-white achievement gap narrowed significantly for cohorts betwen 1950 and 1970.}.

@@

The reasons the authors find a reversal in the trend could be numerous and should be studied very closely. The authors incorporate a number of country-level indicators to try and explaind this change and suggest that the reversal is likely due to the high increase of preschool enrollment. They suggest similarly that in this same period (1998 - 2010) the income achievement gap in early schooling enrollment decreased substantially. Their conclusions are very speculative and have no empirical support, which is why this is still an open question.

@@

One limitation of their study is that they concentrate on countries which have a very particular educational structure, namely the fact that there is little formal stratification in terms of curricula. These four countries have no major jump in school selection, quite the opposite to the average European country. A thorough review by \citet{werfhorst_mijs} sheds some light on the subject. First and foremost, they gather substantive evidence showing that countries which have a highly tracked curriculum tend to have high levels of inequality, measured in terms of achievement gaps. \citet{hanushek_woesmann_tracking}, use the Progress in International Reading Literacy Study (PIRLS), the Trends in International Mathematics and Science Study (TIMSS) and the Programme for International Student Assessment (PISA) surveys to gauge whether highly tracked countries do indeed increase inequality after students pass the age at first selection of tracking. The results suggest that early tracking increases educational inequality. While less clear, there is also a tendency for early tracking to reduce mean performance. \citet{micklewright} using PISA but a different empirical strategy, find that countries which have a high level of tracking, are distinctively unequal in the difference between the top 95th and bottom 5th performers. In fact, the difference in test scores between these two groups is about 10 times higher then the average annual gain of a year of schooling.

@@

Second, we will be able, for the first time, to study the evolution of the achievement gap for several countries other than the United States. This will help asses the mangitude of the inequality as well as the rate of growth/decline. This study not only documents the size and changes of the SES gap, but pays particular attention to the source of the changes. That is, we study whether bottom performers are falling behind, if top performers are gaining advantages, or if both phenomenas are simultaneously at play.

More formally, We're interested in studying
a) Which countries are experiencing changes in the achievement gap;
b) the rate at which each country-gap is widening/narrowing and;
c) establishing whether the gaps are widening/narrowing because particular groups are getting ahead/behind.
d) The size of the achievement in a comparative perspective and its relationship to overall achievement levels;

@@

a) The seminal work of \citet{reardon2011} suggests that achievement gaps change, and they do so much quicker that we though. After recording a SES gap increase of about 40\% in only 30 years, \citep{reardon_portilla} stress that they also found a significant decrease in only 15 years of data. This shows that it is important to study the changes in the achievement gap over time. This will bring a useful comparison and pinpoint which countries are doing the best to reduce the gap.

@@

c) The widening/narrowing of the achievement gap has a source, which has been often studied to be related to everything from educational spending, income inequality, time allocation to students and preschool enrollment. The literature has concentrated very narrowly on whether the gap is increasing because the top performers are getting ahead, because both are distancing or because the bottom is falling behind. We shall pay particular attention to identifying the rate at which the top/bottom groups are evolving over time. This type of analysis is particularly useful in jotting down the mechanisms through which the gap is evolving. If the bottom performers are stagnated whereas the top is gaining ground, the efficiency-trade off dilemma of tracking might be much more credible than other explanations.

@@

PISA collects data through a two-stage stratified sampling design. With the help of official governments, PISA randomly chooses 150 schools in each country, where they then randomly pick thirty 15 year olds to undertake the two hour tests. The sample size for PISA 2000 is \Sexpr{country_rows[1]}, \Sexpr{country_rows[2]} for PISA 2003, \Sexpr{country_rows[3]} for PISA 2006, \Sexpr{country_rows[4]} for PISA 2009, \Sexpr{country_rows[5]} for PISA 2012, and \Sexpr{country_rows[6]} for PISA 2015. Together with the subject tests, PISA collects personal information from students, their families and their school environment (including teacher surveys), that serves as relevent background information that can be matched to the students performance. With the recent inclusion of PISA 2015, these six waves make up a time-series analysis of 15 years, enough to visualize changes in the structure of an educational system. None of the literature cited so far has used the last PISA wave, which was recently released in December 2016.

@@

Another social background variable is the International Socio-Economic Index of Occupational Status (ISEI). This variable attempts to capture the social status of the family, without asking for income information. This index variable was developed by \citet{ganzeboom1996, ganzeboom2010} and it attempts to measure occupational status using a continuous measure. This indicator is a very reliable alternative to the  classical Erikson-Goldthorpe-Portocarero classification \citet{erikson1979}. It has been scaled for comparability between waves and some authors have used it for inequality studies, finding expected results to be consistent with the social mobility literature \citep{anna2016_global}. PISA also includes a plethora of indicators on family wealth, home educational resources, the number of books in the home, among many other material resources in the household.

@@

@@

For each country in each wave, SES disparities in achievement are measured as the gap in standardized achievement between the 90th and 10th percentiles of each country’s distribution of each SES variable, following Reardon’s (2011) method for income achievement gaps. The original strategy of \citet{reardon2011} is as follows: first, achievement is standardized (see below for a mathematical explanation of the standardization). We then use it to calculate the mean achievement (and standard error) for each category of the SES variable of interest (parent's education, income categories, etc..). "Category means are plotted at their percentile ranks and cubic models are fit through the points using weighted least squares. Finally, achievement at each country’s 90th and 10th SES percentiles is interpolated from the model" \citep{anna2016_global}. The result is a SES gap from an ordinal variable of interest.

@@

\begin{equation}
Y_i = \alpha + \beta_1 * AGE_i + \epsilon_i, \quad \epsilon_i ~ N(0, \sigma^2)
\end{equation}

@@

@@

\begin{figure}
\begin{center}
<<variance_pisa, include = T, out.height = '3in', out.width = '3.5in', fig.align = 'center'>>=

@@

@@

@@

\begin{equation}
Y_i = \alpha_{j[i]} + \beta_{j[i]} * SES_i + \epsilon_i,\ \text{for} \ i \ \text{= 1, 2, ...,} \ n \ \text{within country} \ j
\end{equation}

@@

We also calculate the standard error of this difference and use it to create uncertainty intervals in the estimation of time trends. We fit a multilevel rather than a linear model because by pooling the information together, we weight countries appropriately to their sample size. Given that including the SES dummy reduces the sample size considerably, we want to be able to estimate each country-difference as accurately as possible. For further exploration, we also compute the 90th/50th and 50th/10th SES gaps following the same method outlined above.

>>>>

<<default_conf>>=
  library(knitr)
  library(arm)
  library(saves)
  library(haven)
  library(PISA2000lite)
  library(PISA2003lite)
  library(PISA2006lite)
  library(PISA2009lite)
  library(PISA2012lite)
  library(intsvy)
  library(cimentadaj) # devtools::install_github("cimentadaj/cimentadaj")
  library(countrycode) # For region variable
  library(car)
  library(SAScii)
  library(inequalityintsvy) # devtools::install_github("cimentadaj/inequalityintsvy")
  library(lme4)
  library(modelr)
  library(tidyverse)
  library(ggrepel)
  library(xtable)
  library(ggthemes)
  library(matrixStats)

@@

      # Calculate median math score of all PV's
      .x$dv <- apply(.x[test_vars], 1, median, na.rm = T)
      .x$dv <- rowMedians(as.matrix(.x[test_vars]), na.rm = T)

@@

\section{Literature Review}
% You have a bunch of citations to take from here
Educational inequality and its long term impacts is a topic that has been prominent in the social science literature for the past 30 years. The idea of meritocracy and intergenerational transfers have motivated, in conservative terms, much of the research on social mobility and social inequalities, for a good part of the 20th century and 21st century. When James Coleman released his famous Coleman report \citep{coleman1966}, it raised the topic to starlight by suggesting that family SES and student performance are tightly linked. The topic has been studied extensively since the report was released and several authors have contested whether the relationship is an invariable social law or a product of institutional arrangements. For a detailed review of the long literature on educational inequality, please see \citet{gamoran2001}. But fast-forward until today, we've understood the relationship much better.
Educational inequality and its long term impacts is a topic that has been prominent in the social science literature for the past 30 years. The idea of meritocracy has motivated, in conservative terms, much of the research on social mobility and social inequalities, for a good part of the 20th century and 21st century. When James Coleman released his famous Coleman report \citep{coleman1966}, it raised the topic to starlight by suggesting that family SES and student performance are tightly linked. The topic has been studied extensively since the report was released and several authors have contested whether the relationship is an invariable social law or a product of institutional arrangements. For a detailed review of the long literature on educational inequality, please see \citet{gamoran2001}. But fast-forward until today, we've understood the relationship much better.

The work of James Heckman, for example, has clearly showed that cognitive and non-cognitive inequalities are present before entering school \citep{heckman2006}. This means that inequality starts strong before the schooling of a child begins and schools can do so much to close it. These inequalities can be narrowed significantly with high quality learning environments, specially for children coming from unfavourable conditions. But one of the most important findings from \citet{cunha2006} is that the cognitive level of a child in time \(t\) is a direct function of the experiencies from time \(t-1\). Although this sounds logical, it's importance is not understood entirely.
% Here, instead of James Heckman, you should cite the literature on developmental psychology
% which had already discovered much of what heckman was saying.

The implications of this cumulative model suggests that it is much less cost-effective to invest in children in time \(t\) that in \(t-1\). That is why the whole debate about narrowing inequality between high and low SES families should start in early education -- it is the cheaper than investing in later education and it brings about the largest returns. Despite all of these findings and investments to reduce the gaps, we still find that the relationship between parental education or socio-economic status is present in virtually all empirical studies of social mobility and inequalities \citep{breen1997, breen2007, waldfogel2006, bradbury2015, chetty2016}. What we don't know is how and why this relationship is formed from such an early age. Moreover, many have suggested different explanations and the topic has been very contested \citep{brooks1997}.
% And schools can do a great deal to diminish it
The work of James Heckman, for example, has clearly showed that cognitive and non-cognitive inequalities are present before entering school \citep{heckman2006}. This means that inequality starts strong before the schooling of a child begins and schools can help diminish the influence of the origin of a child. The impact of inequality can be eased significantly if children are enrolled in high quality learning environments, specially for children coming from unfavourable conditions. But one of the most important findings from \citet{cunha2006} is that the cognitive level of a child in time \(t\) is a direct function of the experiencies from time \(t-1\). Although this sounds logical, it's importance is not understood entirely.

The implications of this cumulative model suggests that it is much less cost-effective to invest in children in time \(t\) that in \(t-1\). That is why the whole debate about narrowing inequality between high and low SES families should start in early education -- it is the cheaper than investing in later education and it brings about the largest returns. Despite all of these findings and investments to reduce the gaps, we still find that the relationship between parental education or socio-economic status is present in virtually all empirical studies of social mobility and inequalities \citep{breen1997, breen2007, waldfogel2006, bradbury2015, chetty2016}.

@@

From Denmark, for example, the literature has learned a great deal about how to improve social mobility. But that was possible by first learning that Denmark is by far one of the most mobile countries, boosting the chances of upward mobility specially for less advantaged children \citep{bjorklund2009, jaeger2007}. The important finding came when we discovered the main reason behind their social escalator: it's educational system. For example, research by \citet{esping_waldfogel2012} and \citet{bauchmuller2014}, show that the Danish early education system has important and longstanding impacts in helping children reach their fullest potential. It does so very strongly for children which otherwise would be doomed to have a bleak future. Moreover, it tracks students into different curricula very late compared to other Europen countries (age 16), something which has been linked to less educational inequality. These two traits make the Danish system uniquely effective. The early schooling experience attempts to level children to the same degree and this process is not stopped by different curricular tracks because it starts at around age 16 when cognitive abilities are less malleable. In short, the importance of this finding, is that we should step back and first study whether the effect is there, how big is it, and then proceed to find the causes behind it.
From Denmark, for example, the literature has learned a great deal about how to improve social mobility. But that was possible by first learning that Denmark is by far one of the most mobile countries, boosting the chances of upward mobility specially for less advantaged children \citep{bjorklund2009, jaeger2007}. The important finding came when we discovered the main reason behind their social escalator: it's educational system. For example, research by \citet{esping_waldfogel2012} and \citet{bauchmuller2014}, show that the Danish early education system has important and longstanding impacts in helping children reach their fullest potential. It does so very strongly for children which otherwise would be doomed to have a bleak future. Moreover, it tracks students into different curricula very late compared to other Europen countries (age 16), something which has been linked to less educational inequality. These two traits make the Danish system uniquely effective. Preschool education attempts to level children to the same degree and this process is not stopped by different curricular tracks because it starts at around age 16 when cognitive abilities are less malleable. In short, the importance of this finding, is that we should step back and first study whether the effect is there, how big is it, and then proceed to find the causes behind it.

The first attempt to study the evolution of the achievement gap has found that the gap in cognitive abilities between High-SES and Low-SES kids has been widening over the years. The literature on the topic has mainly concentrated on studying the American case \citep{reardon2011} but other international evidence is emerging with a very similar landscape. The United States is usually the case of study because it is the only country where cognitive testing is present as early as 1940. This time-trend has allowed researchers to study achievement gaps in a lengthy period of time. Using this information, \citet{reardon2011} is the first to investigate the evolution of the cognitive gap and the results are very surprising. Not only has the cognitive gap between the 90th income percentile and the 10th income percentile grown over time, but it has grown faster and to be wider than the highly contested white-black gap \citep{magnuson2008}. The gaps have actually reversed and now a days we find that the income achievement gap is nearly twice as large as the black-white achievement gap (quite the opposite to 20 years back).
% widening over the years? Gosta, page 3
% income achievement gap, gosta page 3

\citet{reardon2011} finds that the increase in the gaps has occured predominantely from the 1970's until the 2000's. In fact, the hard numbers suggest that the gap widened by 40-50\%. The author also estimates the rate of change using data as early as 1940 and finds an even higher increase of   75\%. Given that the studies before 1970 are less reliable in terms of comparability and sampling design, the author computes all results for both before/after 1970. To provide a definitive answer to the size of the gap, \citet{reardon2011} concludes that the U.S gap between the 90th and 10th income percentile is at about 1.25 standard deviations in the year 2001. Using longitudinal data, \citet{bradbury2015} find similar results. Their empirical analysis suggests that for American 14 year olds, the gap is above 1 standard deviation but lower than 1.25. Surprisingly, \citet{duncan2011} find similarly to the previous studies and confirm a gap of  1.25-1.50 standard deviations. One important drawback of these studies is that they don't present the uncertainty of this estimates. Not necessarily to gauge their statistical significance, but to simply asses how much we can trust their accuracy. It could be that the gap is at  1.25 standard deviations, varying up to 1.75 and down to 0.80. Right now we don't know the upper/lower bound of this estimate, making it difficult to compare when new evidence arises.
The first attempt to study the evolution of the achievement gap has found that the gap in cognitive abilities between High-SES and Low-SES kids has been widening over the years. The literature on the topic has mainly concentrated on studying the American case \citep{reardon2011} but other international evidence \citep{anna2016_global} is emerging with a very similar landscape. The United States is usually the case of study because it is the only country where cognitive testing is present as early as 1940. This time-trend has allowed researchers to study achievement gaps in a lengthy period of time. Using this information, \citet{reardon2011} is the first to investigate the evolution of the cognitive gap and the results are very surprising. Not only has the cognitive gap between the 90th income percentile and the 10th income percentile grown over time, but it has grown faster and to be wider than the highly contested white-black gap \citep{magnuson2008}. The gaps have actually reversed and now a days we find that the income achievement gap is nearly twice as large as the black-white achievement gap (quite the opposite to 20 years back).

Interestingly, this widening of the achievement gap has been paralleled by a growth of income inequality, a very suggestive explanation. However likely, this is a highly contested topic. A recent \href{http://www.epi.org/files/pdf/101972.pdf}{report} by the Economic Policy Institute finds that the black-white income gap has been growing since 1979, very similar to the increase of income inequality. However, if the achiement gap would be a function of income inequality, we should expect for the white-black achievement gap to be widening rather than narrowing, and those are not the findings given by \citet{reardon2011} and \citet{magnuson2008}.
\citet{reardon2011} finds that the increase in the gaps has occured predominantely from the 1970's until the 2000's. In fact, the hard numbers suggest that the gap widened by 40-50\%. The author also estimates the rate of change using data as early as 1940 and finds an even higher increase of   75\%. Given that the studies before 1970 are less reliable in terms of comparability and sampling design, the author computes all results for both before/after 1970. To provide a definitive answer to the size of the gap, \citet{reardon2011} concludes that the U.S gap between the 90th and 10th income percentile is at about 1.25 standard deviations in the year 2001. Using longitudinal data, \citet{bradbury2015} find similar results. Their empirical analysis suggests that for American 14 year olds, the gap is above 1 standard deviation but lower than 1.25. Surprisingly, \citet{duncan2011} find similarly to the previous studies and confirm a gap of  1.25-1.50 standard deviations. One important drawback of these studies is that they don't present the uncertainty of this estimates. Not necessarily to gauge their statistical significance, but to simply asses how much we can trust their accuracy. It could be that the gap is at  1.25 standard deviations, varying up to 1.75 and down to 0.80. Right now we don't know the upper/lower bound of this estimate, making it difficult to compare when new evidence arises.

\footnote{The black-white achievement gap narrowed significantly for cohorts betwen 1950 and 1970.}.
Interestingly, this widening of the achievement gap has been paralleled by a growth of income inequality, a very suggestive explanation. However likely, this is a highly contested topic. A recent \href{http://www.epi.org/files/pdf/101972.pdf}{report} by the Economic Policy Institute finds that the black-white income gap has been growing since 1979, very similar to the increase of income inequality. However, if the achiement gap would be a function of income inequality, we should expect for the white-black achievement gap to be widening rather than narrowing, and those are not the findings given by \citet{reardon2011} and \citet{magnuson2008}\footnote{The black-white achievement gap narrowed significantly for cohorts betwen 1950 and 1970.}.
.
% explain more in detail white income inequality is linked to to the achievement gap
% gosta page 3

@@

The reasons the authors find a reversal in the trend could be numerous and should be studied very closely. The authors incorporate a number of country-level indicators to try and explaind this change and suggest that the reversal is likely due to the high increase of preschool enrollment. They suggest similarly that in this same period (1998 - 2010) the income achievement gap in early schooling enrollment decreased substantially. Their conclusions are very speculative and have no empirical support, which is why this is still an open question.
The reasons the authors find a reversal in the trend could be numerous and should be studied very closely. The authors incorporate a number of country-level indicators to try explain this change and suggest that the reversal is likely due to the high increase of preschool enrollment. In fact, the point out that in this same period (1998 - 2010) the SES gap in preschool enrollment decreased substantially. Their conclusions are very speculative and have no empirical support, which is why this is still an open question.

%

@@

One limitation of their study is that they concentrate on countries which have a very particular educational structure, namely the fact that there is little formal stratification in terms of curricula. These four countries have no major jump in school selection, quite the opposite to the average European country. A thorough review by \citet{werfhorst_mijs} sheds some light on the subject. First and foremost, they gather substantive evidence showing that countries which have a highly tracked curriculum tend to have high levels of inequality, measured in terms of achievement gaps. \citet{hanushek_woesmann_tracking}, use the Progress in International Reading Literacy Study (PIRLS), the Trends in International Mathematics and Science Study (TIMSS) and the Programme for International Student Assessment (PISA) surveys to gauge whether highly tracked countries do indeed increase inequality after students pass the age at first selection of tracking. The results suggest that early tracking increases educational inequality. While less clear, there is also a tendency for early tracking to reduce mean performance. \citet{micklewright} using PISA but a different empirical strategy, find that countries which have a high level of tracking, are distinctively unequal in the difference between the top 95th and bottom 5th performers. In fact, the difference in test scores between these two groups is about 10 times higher then the average annual gain of a year of schooling.
One limitation of their study is that they concentrate on countries which have a very particular educational structure, namely the fact that there is little curricular tracking. These four countries have no major jump in school selection, quite the opposite to the average European country. A thorough review by \citet{werfhorst_mijs} sheds some light on the subject. First and foremost, they gather substantive evidence showing that countries which have a highly tracked curriculum tend to have high levels of inequality, measured in terms of achievement gaps. \citet{hanushek_woesmann_tracking}, use the Progress in International Reading Literacy Study (PIRLS), the Trends in International Mathematics and Science Study (TIMSS) and the Programme for International Student Assessment (PISA) surveys to gauge whether highly tracked countries do indeed increase inequality after students pass the age at first selection of tracking. The results suggest that early tracking increases educational inequality. While less clear, there is also a tendency for early tracking to reduce mean performance. \citet{micklewright} using PISA but a different empirical strategy, find that countries which have a high level of tracking, are distinctively unequal in the difference between the top 95th and bottom 5th performers. In fact, the difference in test scores between these two groups is about 10 times higher then the average annual gain of a year of schooling.

@@

Second, we will be able, for the first time, to study the evolution of the achievement gap for several countries other than the United States. This will help asses the mangitude of the inequality as well as the rate of growth/decline. This study not only documents the size and changes of the SES gap, but pays particular attention to the source of the changes. That is, we study whether bottom performers are falling behind, if top performers are gaining advantages, or if both phenomenas are simultaneously at play.
Second, we will be able, for the first time, to study the evolution of the achievement gap for several countries other than the United States. We will do this by studying the changes in the achievement gap in a time-frame of 15 years. This will help asses the magnitude of the inequality as well as the rate of growth/decline. This study not only documents the size and changes of the SES gap, but pays particular attention to the source of the changes. That is, we study whether bottom performers are falling behind, if top performers are gaining advantages, or if both phenomenas are simultaneously at play.

More formally, We're interested in studying
More formally, we're interested in studying
a) Which countries are experiencing changes in the achievement gap;
b) the rate at which each country-gap is widening/narrowing and;
c) establishing whether the gaps are widening/narrowing because particular groups are getting ahead/behind.
d) The size of the achievement in a comparative perspective and its relationship to overall achievement levels;

@@

a) The seminal work of \citet{reardon2011} suggests that achievement gaps change, and they do so much quicker that we though. After recording a SES gap increase of about 40\% in only 30 years, \citep{reardon_portilla} stress that they also found a significant decrease in only 15 years of data. This shows that it is important to study the changes in the achievement gap over time. This will bring a useful comparison and pinpoint which countries are doing the best to reduce the gap.
a) The seminal work of \citet{reardon2011} suggests that achievement gaps change, and they do so much quicker that we though. After recording a SES gap increase of about 40\% in only 30 years, \citet{reardon_portilla} stress that they also found a significant decrease in only 15 years of data. This shows that it is important to study the changes in the achievement gap over time. This will bring a useful comparison and pinpoint which countries are doing the best to reduce the gap.

@@

c) The widening/narrowing of the achievement gap has a source, which has been often studied to be related to everything from educational spending, income inequality, time allocation to students and preschool enrollment. The literature has concentrated very narrowly on whether the gap is increasing because the top performers are getting ahead, because both are distancing or because the bottom is falling behind. We shall pay particular attention to identifying the rate at which the top/bottom groups are evolving over time. This type of analysis is particularly useful in jotting down the mechanisms through which the gap is evolving. If the bottom performers are stagnated whereas the top is gaining ground, the efficiency-trade off dilemma of tracking might be much more credible than other explanations.
c) The widening/narrowing of the achievement gap has a source, which has been often studied to be related to everything from educational spending, income inequality, time allocation to students and preschool enrollment. The literature has concentrated very narrowly on whether the gap is increasing because the top performers are getting ahead, because both are distancing or because the bottom is falling behind. We shall pay particular attention to identifying the rate at which the top/bottom groups are evolving over time. This type of analysis is particularly useful in jotting down the mechanisms through which the gap is evolving. For example, if whether bottom performers are stagnated and the top is gaining ground, or is it the other way around. Or is it the case that both SES groups are distancing themselves at the same rate.

@@

PISA collects data through a two-stage stratified sampling design. With the help of official governments, PISA randomly chooses 150 schools in each country, where they then randomly pick thirty 15 year olds to undertake the two hour tests. The sample size for PISA 2000 is \Sexpr{country_rows[1]}, \Sexpr{country_rows[2]} for PISA 2003, \Sexpr{country_rows[3]} for PISA 2006, \Sexpr{country_rows[4]} for PISA 2009, \Sexpr{country_rows[5]} for PISA 2012, and \Sexpr{country_rows[6]} for PISA 2015. Together with the subject tests, PISA collects personal information from students, their families and their school environment (including teacher surveys), that serves as relevent background information that can be matched to the students performance. With the recent inclusion of PISA 2015, these six waves make up a time-series analysis of 15 years, enough to visualize changes in the structure of an educational system. None of the literature cited so far has used the last PISA wave, which was recently released in December 2016.
PISA collects data through a two-stage stratified sampling design. With the help of official governments, PISA randomly chooses 150 schools in each country, where they then randomly pick thirty 15 year olds to undertake the two hour tests. The sample size for PISA 2000 is \Sexpr{country_rows[1]}, \Sexpr{country_rows[2]} for PISA 2003, \Sexpr{country_rows[3]} for PISA 2006, \Sexpr{country_rows[4]} for PISA 2009, \Sexpr{country_rows[5]} for PISA 2012, and \Sexpr{country_rows[6]} for PISA 2015. Together with the subject tests, PISA collects personal information from students, their families and their school environment (including teacher surveys), that serves as relevent background information that can be matched to the students performance. With the recent inclusion of PISA 2015, we have a new opportunity to study six waves that make up a time-series analysis of 15 years. None of the literature cited so far has used the last PISA wave, which was recently released in December 2016.

@@

Another social background variable is the International Socio-Economic Index of Occupational Status (ISEI). This variable attempts to capture the social status of the family, without asking for income information. This index variable was developed by \citet{ganzeboom1996, ganzeboom2010} and it attempts to measure occupational status using a continuous measure. This indicator is a very reliable alternative to the  classical Erikson-Goldthorpe-Portocarero classification \citet{erikson1979}. It has been scaled for comparability between waves and some authors have used it for inequality studies, finding expected results to be consistent with the social mobility literature \citep{anna2016_global}. PISA also includes a plethora of indicators on family wealth, home educational resources, the number of books in the home, among many other material resources in the household.
Another social background variable is the International Socio-Economic Index of Occupational Status (ISEI). This variable attempts to capture the social status of the family, without asking for income information. This index variable was developed by \citet{ganzeboom1996, ganzeboom2010} and it attempts to measure occupational status using a continuous measure. This indicator is a very reliable alternative to the  classical Erikson-Goldthorpe-Portocarero classification \citep{erikson1979}. It has been scaled for comparability between waves and some authors have used it for inequality studies, finding expected results to be consistent with the social mobility literature \citep{anna2016_global}. PISA also includes a plethora of indicators on family wealth, home educational resources, the number of books in the home, among many other material resources in the household.

@@

% Talk in more detail about the ESCS index - is it a factor analysis? What's the methodology like?

@@

% We then use it to calculate the mean achievement and (standard error or standard deviation)?
For each country in each wave, SES disparities in achievement are measured as the gap in standardized achievement between the 90th and 10th percentiles of each country’s distribution of each SES variable, following Reardon’s (2011) method for income achievement gaps. The original strategy of \citet{reardon2011} is as follows: first, achievement is standardized (see below for a mathematical explanation of the standardization). We then use it to calculate the mean achievement (and standard error) for each category of the SES variable of interest (parent's education, income categories, etc..). "Category means are plotted at their percentile ranks and cubic models are fit through the points using weighted least squares. Finally, achievement at each country’s 90th and 10th SES percentiles is interpolated from the model" \citep{anna2016_global}. The result is a SES gap from an ordinal variable of interest.

@@

% Justify why you use the AGE variable
\begin{equation}
Y_i = \alpha + \beta_1 * AGE_i + \epsilon_i, \quad \epsilon_i ~ N(0, \sigma^2)
\end{equation}

@@

% Change the hat for something else because it makes the notation confusing.

@@

% Try plotting this for each country and wave
% Like, every country as a row and 6 columns for each wave.
\begin{figure}
\begin{center}
<<variance_pisa, include = T, out.height = '3in', out.width = '3.5in', fig.align = 'center'>>=

@@

% As well, updated y hat inside the equation with the updated y from the
% equation above. And then change the hat from the left hand side with a new indicator.

@@

% Explain more in detail how the reliability indicator is calculated.

@@

% Replace the Y_i with the chosen y from the previos equation.
\begin{equation}
Y_i = \alpha_{j[i]} + \beta_{j[i]} * SES_i + \epsilon_i,\ \text{for} \ i \ \text{= 1, 2, ...,} \ n \ \text{within country} \ j
Y_i = \alpha_{j[i]} + \beta_{j[i]} * SES_i + \epsilon_i,\ \text{for} \ i \ \text{= 1, 2, ...,} \ n \ \text{for each country} \ j
\end{equation}

% Try running also a linear model together with the multilevel model and ompcare results
% Brunno suggested this because countries which might have a very different SES estimate
% might be moved more towards the average.

@@

We also calculate the standard error of this difference and use it to create uncertainty intervals in the estimation of time trends. We fit a multilevel rather than a linear model because by pooling the information together, we weight countries appropriately to their sample size. Given that including the SES dummy reduces the sample size considerably, we want to be able to estimate each country-difference as accurately as possible. For further exploration, we also compute the 90th/50th and 50th/10th SES gaps following the same method outlined above.
We also calculate the standard error of this difference and use it to create uncertainty intervals in the estimation of time trends. We fit a multilevel rather than an OLS model because by pooling the information together, we weight countries appropriately to their sample size. Given that including the SES dummy reduces the sample size considerably, we want to be able to estimate each country-difference as accurately as possible. For further exploration, we also compute the 90th/50th and 50th/10th SES gaps following the same method outlined above.
% Explain more in detail that it's a dummy and not three groups.
